Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-22 paper/ms.tex
update ms

% upon AAS submission
%\documentclass[12pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
\turnoffedit

@@

		\vspace{-0.65cm}
		\subfloat{
			\includegraphics[width=0.66\textwidth]{f1b.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Complex periodic variables (CPVs)}:
		{\it Top:} Phase-folded TESS light curves for three CPVs.  Each
		panel shows the average of the data accumulated over one month,
		relative to the mean stellar brightness.
		Gray circles are raw 2-minute data; black circles are binned to
		300 points per cycle.  The period in hours is printed in the
		bottom right corner.  Left-to-right, the objects are LP 12-502
    (TIC~402980664; Sector~19), TIC~94088626 (Sector 10), and
    TIC~425933644 (Sector~28).
    {\it Bottom:} Cartoon configurations for magnetically-entrained
    corotating material.  The dust clump scenario (left) and gas
    prominence scenario (right) propose different opacity sources, and
    different occultation geometries.
	}
	\label{fig:f1}
\end{figure*}

@@

We performed an independent search using a Fourier-based approach,
following \citet{2019ApJ...876..127Z} and \citet[][their
Section~1.3]{2023MNRAS.524.4220P}.  Starting with the {\tt PDC\_SAP}
light curves, we normalized each light curve, and then re-binned it
into equal width 2-minute bins to account for the uneven spacing in
the TESS data, as well as the data gap caused by satellite downlink
during each sector.  We then padded the data to ensure that the light
curve had a length that was a power of two, as described by
\citeauthor{2019ApJ...876..127Z}.  After taking the Fourier transform
of the padded light curve using {\tt numpy.fft}, we searched for peaks
with a significance exceeding 12-$\sigma$ within a set of 500
frequency bins.

@@

\subsubsection{Ages}
While most of our target stars were field stars, color--absolute
magnitude diagrams suggested that most CPVs tended to be on the
pre-main-sequence (e.g.\ panel {\it (k) of Figure~\ref{fig:vet}}).  We
therefore estimated stellar ages by checking for probabilistic spatial
and kinematic associations between the CPVs and known clusters in the
solar neighborhood.  For most stars in our sample, we did this using
BANYAN\,$\Sigma$
\citep{2018ApJ...856...23G}.\footnote{\url{https://github.com/jgagneastro/banyan_sigma},
git commit \texttt{394b486}} This algorithm calculates the probability
that a given star belongs to either the field, or to any of 27 young
clusters (``associations'') within 150\,pc of the Sun.  This is
achieved by modeling the field and cluster populations as multivariate
Gaussian distributions in 3-D position and 3-D velocity space.  We
used the Gaia DR2 sky positions, proper motions, and distances to
calculate the membership probabilities.  BANYAN\,$\Sigma$ in turn
analytically marginalizes over the radial velocity dimension.  The
probabilities returned by this procedure are qualitatively helpful,
but should be interpreted with caution because the assumption of
Gaussian distributions is questionable for most groups within the
solar neighborhood \citep[see e.g.][Figure~10]{2021ApJ...917...23K}.

@@

{\it Gaia DR3 non-single stars}---Gaia DR3 included a {\tt
non\_single\_star} column that flagged eclipsing, astrometric, and
spectroscopic binaries.  None of the stars in our CPV sample were
identified as possible binaries in this column.

@@

The groups that contain the largest number of CPVs in our catalog are
Sco-Cen, Tuc-Hor, and Columba.  Six CPVs were also identified in the
Argus association \citep{2019ApJ...870...27Z}, which serves as an
indirect line of evidence supporting the reality and youth of that
group.  The large contribution from Sco-Cen is not surprising since
Sco-Cen contains the majority of pre-main-sequence stars in the solar
neighborhood, and many of its stars were selected for TESS 2-minute
cadence observations by guest investigators.  Given the
$\lesssim$$10\%$ completeness of TESS beyond 100\,pc
(Figure~\ref{fig:completeness}), there may be many more CPVs in
Sco-Cen that remain to be discovered.

@@

Figure~\ref{fig:evoln} shows ``before'' and ``after'' views of 27 CPVs
for which TESS 2-minute cadence observations were available at least
two years apart.  Such a baseline was available for 32 of the \ngoods\
confirmed CPVs in our catalog; for plotting purposes we show the
brightest 27.  We have defined $\phi=0$ for each sector to be the time
of minimum light observed in that sector, rather than using a
consistent phase definition across multiple sectors.  This is because
for most of the sources we do not know the period at the precision
necessary to be able to accurately propagate an ephemeris over two
years.  The achievable period precision, $\sigma_P$, can be estimated
as
\begin{equation}
  \sigma_{P} = \frac { \sigma_{\phi} P } { N_{\rm baseline} },
  \label{eq:periodprecision}
\end{equation}
for $N_{\rm baseline}$ the number of cycles in the observed baseline
and $\sigma_{\phi}$ the phase precision with which any one feature
(e.g.~a dip, or the overall shape of the sinusoidal envelope) can be
tracked.  Assuming $\sigma_\phi$$\approx$$0.02$ and a 20-day baseline
over a single TESS sector yields
$\sigma_{P}$$\approx$$0.25^{+0.38}_{-0.14}$\,minutes for the
population shown in Figure~\ref{fig:evoln}; propagated forward 1{,}000
cycles yields a typical ephemeris uncertainty range of 2-11\,hours.
Measuring the period independently for each sector did not reveal
evidence for significant ($>$3$\sigma$) changes in period, implying a
period stability of $\lesssim$0.1\% over two years.

@@

The observed lifetime of the CPV phenomenon could provide another way
to discern between the gas and dust clump scenarios.  Based on the
available statistics from e.g.  \citet{2022AJ....164...80R} and
references therein, it seems plausible that CPV occurrence decreases
with stellar age from $\approx$3\% at 10\,Myr (Sco-Cen), to
$\approx$1\% at 100\,Myr (Pleiades), down to 0\% by the
$\approx$700\,Myr age of Praesepe.  This is odd in the context of the
prominence scenario, because pre-main-sequence M dwarfs spin up over
the first 100\,Myr; prominences might therefore be expected to be {\it
more} common at 100\,Myr than at 10\,Myr, under the assumption that
the production of prominences depends only on the stellar rotation
rate.  The dust clump scenario would hold a natural explanation: the
lower occurrence of CPVs around older stars would simply reflect a
finite supply of dust.  One possible complication however is that the
magnetic field topology of rapidly rotating M dwarfs may depend on
factors other than the rotation rate (e.g.\ the age), which might
alter the production of prominences.

>>>>

% upon AAS submission
%\documentclass[12pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

		\vspace{-0.62cm}
		\subfloat{
			\includegraphics[width=0.66\textwidth]{f1b.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Complex periodic variables (CPVs)}:
		{\it Top:} Phase-folded TESS light curves for three CPVs.  Each
		panel shows the average of the data accumulated over one month,
		relative to the mean stellar brightness.
		Gray circles are raw 2-minute data; black circles are binned to
		300 points per cycle.  The period in hours is printed in the
		bottom right corner.  Left-to-right, the objects are LP 12-502
    (TIC~402980664; Sector~19), TIC~94088626 (Sector 10), and
    TIC~425933644 (Sector~28).
    {\it Bottom:} Cartoon configurations for magnetically-entrained
    corotating material.  The dust clump scenario (left) and gas
    prominence scenario (right) propose different opacity sources, and
    different occultation geometries.
	}
	\label{fig:f1}
\end{figure*}

@@

We performed an independent search using a Fourier-based approach,
following \citet{2019ApJ...876..127Z} and \citet[][their
Section~1.3]{2023MNRAS.524.4220P}.  Starting with the {\tt PDC\_SAP}
light curves, we normalized each light curve, and then re-binned it
into equal width 2-minute bins to account for the uneven spacing in
the TESS data, as well as the data gap caused by satellite downlink
during each sector.  We then padded the data to ensure that the light
curve had a length that was a power of two, as described by
\citeauthor{2019ApJ...876..127Z}\ After taking the Fourier transform
of the padded light curve using {\tt numpy.fft}, we searched for peaks
with a significance exceeding 12-$\sigma$ within a set of 500
frequency bins.

@@

\subsubsection{Ages}
While most of our target stars were field stars, color--absolute
magnitude diagrams suggested that most CPVs tended to be on the
pre-main-sequence (e.g.\ panel {\it (k)} of Figure~\ref{fig:vet}).  We
therefore estimated stellar ages by checking for probabilistic spatial
and kinematic associations between the CPVs and known clusters in the
solar neighborhood.  For most stars in our sample, we did this using
BANYAN\,$\Sigma$
\citep{2018ApJ...856...23G}.\footnote{\url{https://github.com/jgagneastro/banyan_sigma},
git commit \texttt{394b486}} This algorithm calculates the probability
that a given star belongs to either the field, or to any of 27 young
clusters (``associations'') within 150\,pc of the Sun.  This is
achieved by modeling the field and cluster populations as multivariate
Gaussian distributions in 3-D position and 3-D velocity space.  We
used the Gaia DR2 sky positions, proper motions, and distances to
calculate the membership probabilities.  BANYAN\,$\Sigma$ in turn
analytically marginalizes over the radial velocity dimension.  The
probabilities returned by this procedure are qualitatively helpful,
but should be interpreted with caution because the assumption of
Gaussian distributions is questionable for most groups within the
solar neighborhood \citep[see e.g.][Figure~10]{2021ApJ...917...23K}.

@@

{\it Gaia DR3 non-single stars}---Gaia DR3 included a {\tt
non\_single\_star} column that flagged eclipsing, astrometric, and
spectroscopic binaries.  None of the stars in our CPV sample were
identified as \deleted{possible }binaries in this column.

@@

The groups that contain the largest number of CPVs in our catalog are
Sco-Cen, Tuc-Hor, and Columba.  Six CPVs were also identified in the
Argus association \citep{2019ApJ...870...27Z}, which serves as an
indirect line of evidence supporting the reality and youth of that
group.  The large contribution from Sco-Cen is not surprising since
Sco-Cen contains the majority of pre-main-sequence stars in the solar
neighborhood, and many of its stars were selected for TESS 2-minute
cadence observations by guest investigators.  Given the
$\lesssim$$10\%$ completeness of \replaced{TESS}{our data} beyond 100\,pc
(Figure~\ref{fig:completeness}), there may be many more CPVs in
Sco-Cen that remain to be discovered.

@@

Figure~\ref{fig:evoln} shows ``before'' and ``after'' views of 27 CPVs
for which TESS 2-minute cadence observations were available at least
two years apart.  Such a baseline was available for 32 of the \ngoods\
confirmed CPVs in our catalog; for plotting purposes we show the
brightest 27.  We have defined $\phi=0$ for each sector to be the time
of minimum light observed in that sector, rather than using a
consistent phase definition across multiple sectors.  This is because
for most of the sources we do not know the period at the precision
necessary to be able to accurately propagate an ephemeris over two
years.  The achievable period precision, $\sigma_P$, can be estimated
as
\begin{equation}
  \sigma_{P} = \frac { \sigma_{\phi} P } { N_{\rm baseline} },
  \label{eq:periodprecision}
\end{equation}
for $N_{\rm baseline}$ the number of cycles in the observed baseline
and $\sigma_{\phi}$ the phase precision with which any one feature
(e.g.~a dip, or the overall shape of the sinusoidal envelope) can be
tracked.  Assuming $\sigma_\phi$$\approx$$0.02$ and a 20-day baseline
over a single TESS sector yields
$\sigma_{P}$$\approx$$0.25^{+0.38}_{-0.14}$\,minutes for the
population shown in Figure~\ref{fig:evoln}; propagated forward 1{,}000
cycles yields a typical ephemeris uncertainty range of 2 to 11\,hours.
Measuring the period independently for each sector did not reveal
evidence for significant ($>$3$\sigma$) changes in period, implying a
period stability of $\lesssim$0.1\% over two years.

@@

The observed lifetime of the CPV phenomenon could provide another way
to discern between the gas and dust clump scenarios.  Based on the
available statistics from \deleted{e.g. }\citet{2022AJ....164...80R} and
references therein, it seems plausible that CPV occurrence decreases
with stellar age from $\approx$3\% at 10\,Myr (Sco-Cen), to
$\approx$1\% at 100\,Myr (Pleiades), down to 0\% by the
$\approx$700\,Myr age of Praesepe.  This is odd in the context of the
prominence scenario, because pre-main-sequence M dwarfs spin up over
the first 100\,Myr; prominences might therefore be expected to be {\it
more} common at 100\,Myr than at 10\,Myr, under the assumption that
the production of prominences depends only on the stellar rotation
rate.  The dust clump scenario would hold a natural explanation: the
lower occurrence of CPVs around older stars would simply reflect a
finite supply of dust.  One possible complication however is that the
magnetic field topology of rapidly rotating M dwarfs may depend on
factors other than the rotation rate (e.g.\ the age), which might
alter the production of prominences.
